---
title: "carpe-diem-ukb analysis notebook"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
source("R/utils.R")
source("R/packages.R")
```

## configuration

```{r config}
# ── paths ─────────────────────────────────────────────────────────────────────
METADATA_CSV       <- "inputs/ukb_metadata.csv"
DATA_PARTICIPANT   <- "inputs/data_participant.csv"
DATA_PRS           <- "inputs/PRS_pred_allchr.sscore"
DATA_ICD10         <- "inputs/data_diag_hesin.csv"
DATA_PROTEIN_BATCH <- "inputs/pro_batch_fastingtime.rds"
DATA_PROTEIN_MAP   <- "inputs/protein_panel_map.rds"

OUTPUT_FIGURES     <- "figures"
OUTPUT_RESULTS     <- "results"

# ── analysis parameters ───────────────────────────────────────────────────────
CONF_LEVEL <- 0.95
COVARIATES <- c(
  "age", "sex",
  paste0("pc", 1:10)
)

# ── display aesthetics ────────────────────────────────────────────────────────
TRAIT_COLORS <- c(
  "Cardiovascular disease" = "#3B7EA1",
  "Metabolic disease"      = "#B07AA1",
  "Psychiatric disorder"   = "#E39C37",
  "Sociodemographic factor"       = "#2C8C69",
  "Adiposity"              = "#C7B83C",
  "Glycemic"               = "#6BAED6",
  "Cardiovascular" = "#4DAF8A",
  "Lipid"                  = "#C76D2B",
  "Liver"                  = "#7B61A8",
  "Dental caries" = "#E78AC3"
)

PANEL_COLORS <- c(
  "Inflammation" = "#3F7193",
  "Inflammation II" = "#3F7193",
  "Cardiometabolic" = "#9D4764",
  "Cardiometabolic II" = "#9D4764",
  "Oncology" = "#E8A65F",
  "Oncology II" = "#E8A65F",
  "Neurology" = "#449064",
  "Neurology II" = "#449064"
)

FACET_MAP <- c(
  "Cardiovascular disease" = "Diseases & others",
  "Metabolic disease"      = "Diseases & others",
  "Psychiatric disorder"   = "Diseases & others",
  "Sociodemographic factor"= "Diseases & others",
  "Adiposity"              = "Cardiometabolic traits",
  "Glycemic"               = "Cardiometabolic traits",
  "Hormone"                = "Cardiometabolic traits",
  "Lipid"                  = "Cardiometabolic traits",
  "Liver"                  = "Cardiometabolic traits",
  "Cardiovascular" = "Cardiometabolic traits",
  "Dental caries"          = ""
)
```

## load metadata maps

```{r load-maps}
maps <- load_maps(METADATA_CSV)

rename_map       <- maps$rename_map
var_name_map     <- maps$var_name_map
var_trans_map    <- maps$var_trans_map
var_category_map <- maps$var_category_map
var_code_map     <- maps$var_code_map
```

## load and prepare data

```{r load-data}
data_participant <- read.csv(DATA_PARTICIPANT) %>%
  dplyr::rename_with(
    .fn   = ~ unname(rename_map[.x]),
    .cols = dplyr::any_of(names(rename_map))
  ) %>%
  dplyr::mutate(eid = as.numeric(eid))

data_participant$has_dentures <- ifelse(
  data_participant$has_dentures == "Dentures", 1,
  ifelse(data_participant$has_dentures == "None of the above", 0, NA)
)

data_prs <- read.table(DATA_PRS, header = TRUE) %>%
  dplyr::rename(prs = SCORE1_SUM, eid = IID) %>%
  dplyr::mutate(eid = as.numeric(eid))

data_icd10 <- read.csv(DATA_ICD10, stringsAsFactors = FALSE) %>%
  dplyr::mutate(eid = as.numeric(eid))

# merge PRS onto participant data
data_participant <- data_participant %>%
  dplyr::left_join(data_prs, by = "eid")

# join ICD-10 binary flags (data-driven from metadata)
data <- safe_icd_join(
  data         = data_participant,
  data_icd10   = data_icd10,
  metadata_csv = METADATA_CSV,
  id_col       = "eid"
)  %>%
  dplyr::mutate(z_bmi = as.numeric(scale(bmi))) # add bmi for adj for bmi later
```

## define outcomes

```{r define-outcomes}
meta <- maps$metadata

continuous_outcomes <- meta$variable[
  meta$source_type == "direct" &
    !is.na(meta$trait_type) &
    meta$trait_type == "continuous"
]

binary_outcomes <- meta$variable[meta$trait_type == "binary"]

all_outcomes <- c(continuous_outcomes, binary_outcomes)
```

## population characteristics

```{r population-summary}
population_summary <- summarise_population_characteristics_wide(
  df              = data,
  binary_vars     = binary_outcomes,
  continuous_vars = continuous_outcomes,
  sex_col         = "sex",
  age_col         = "age",
  id_col          = "eid"
)
```

## PRS association analyses — without BMI adjustment

```{r prs-associations}
linear_results <- run_lm_grid(
  df           = data,
  outcomes     = continuous_outcomes,
  predictors   = "prs",
  covariates   = COVARIATES,
  idcol        = "eid",
  conf_level   = CONF_LEVEL,
  trait_map    = var_category_map,
  trans_map    = var_trans_map,
  model_filter = "linear"
)

logistic_results <- run_lm_grid(
  df           = data,
  outcomes     = binary_outcomes,
  predictors   = "prs",
  covariates   = COVARIATES,
  idcol        = "eid",
  conf_level   = CONF_LEVEL,
  trait_map    = var_category_map,
  trans_map    = var_trans_map,
  model_filter = "logistic"
)

combined_results <- dplyr::bind_rows(
  linear_results,
  logistic_results,
) %>% dplyr::mutate(pvalue_fdr = stats::p.adjust(pvalue, method = "fdr"))
```

```{r forest-plots, fig.width=10, fig.height=10}
p_combined <- make_forest_plot(
  model_results = combined_results,
  var_map       = var_name_map,
  trait_map     = var_category_map,
  facet_map     = FACET_MAP,
  trait_colors  = TRAIT_COLORS,
  datatype      = "prs_associations",
  plot_type     = "forest_plot",
  outputpath    = OUTPUT_FIGURES,
  width         = 3200,
  height        = 2400
)

p_combined
```

## PRS association analyses — with BMI adjustment

```{r prs-associations-bmi}
linear_results_bmi_adj <- run_lm_grid(
  df           = data,
  outcomes     = setdiff(continuous_outcomes, c("bmi", "waist", "hip", "body_fat_perc")),
  predictors   = "prs",
  covariates   = c(COVARIATES, "z_bmi"),
  idcol        = "eid",
  conf_level   = CONF_LEVEL,
  trait_map    = var_category_map,
  trans_map    = var_trans_map,
  model_filter = "linear"
)

linear_results_bmi_adj_adiposity <- run_lm_grid(
  df           = data,
  outcomes     = c("bmi", "waist", "hip", "body_fat_perc"),
  predictors   = "prs",
  covariates   =  COVARIATES, # here we do not adjust for bmi sds
  idcol        = "eid",
  conf_level   = CONF_LEVEL,
  trait_map    = var_category_map,
  trans_map    = var_trans_map,
  model_filter = "linear"
)

logistic_results_bmi_adj <- run_lm_grid(
  df           = data,
  outcomes     = binary_outcomes,
  predictors   = "prs",
  covariates   = c(COVARIATES, "z_bmi"),
  idcol        = "eid",
  conf_level   = CONF_LEVEL,
  trait_map    = var_category_map,
  trans_map    = var_trans_map,
  model_filter = "logistic"
) 

linear_results_bmi_adj_combined <- dplyr::bind_rows(
  linear_results_bmi_adj,
  linear_results_bmi_adj_adiposity
)

combined_results_bmi_adj <- dplyr::bind_rows(linear_results_bmi_adj_combined, logistic_results_bmi_adj) %>% 
  dplyr::mutate(pvalue_fdr = stats::p.adjust(pvalue, method = "fdr"))
```

```{r, fig.width=16, fig.height=10}
p_combined_bmi_adj <- make_forest_plot(
  model_results = combined_results_bmi_adj,
  var_map       = var_name_map,
  trait_map     = var_category_map,
  facet_map     = FACET_MAP,
  trait_colors  = TRAIT_COLORS,
  datatype      = "prs_associations",
  plot_type     = "forest_plot_bmi_adj",
  outputpath    = OUTPUT_FIGURES,
  width         = 3200,
  height        = 2400
)

p_combined_bmi_adj
```

## Proteomics: data prep

```{r proteomics-setup}
data_protein_batch <- readRDS(DATA_PROTEIN_BATCH) %>%
  dplyr::rename(batch = Batch) %>%
  dplyr::mutate(z_fasting_time = as.numeric(scale(Fasting_time))) %>%
  dplyr::select(-ins_index, -PlateID, -Fasting_time) %>%
  dplyr::rename_with(toupper, -c(eid, batch, z_fasting_time))

protein_panel_map <- readRDS(DATA_PROTEIN_MAP) %>%
  dplyr::rename(protein = Assay, panel = Panel) %>%
  dplyr::mutate(protein = toupper(protein)) %>%
  dplyr::select(protein, panel)

panel_lookup <- stats::setNames(
  protein_panel_map$panel,
  protein_panel_map$protein
)

PROTEINS <- protein_panel_map$protein

data_protein <- data %>%
  dplyr::inner_join(data_protein_batch, by = "eid")
```

# PRS-protein associations - without BMI adjustment

```{r protein-associations}
protein_results <- run_lm_grid(
  df           = data_protein,
  outcomes     = PROTEINS,
  predictors   = "prs",
  covariates   = c(COVARIATES, "z_fasting_time"),
  idcol        = "eid",
  conf_level   = CONF_LEVEL,
  trait_map    = panel_lookup,
  trans_map    = setNames(character(0), character(0)),
  panel_lookup = panel_lookup,
  model_filter = "linear"
) %>% dplyr::mutate(pvalue_fdr = stats::p.adjust(pvalue, method = "fdr"))
```

```{r volcano-plot, fig.width=16, fig.height=4}
p_volcano <- plot_volcano_prs_proteins(
  results_df   = protein_results,
  prs_name_map = c(prs = "PRS"),
  cov_formula  = c(COVARIATES, "z_fasting_time"),
  top_n        = 10,
  p_nominal    = 0.05,
  fdr_cutoff   = 0.05,
  x_step       = 0.2,
  outputpath   = OUTPUT_FIGURES,
  save         = TRUE,
  dpi          = 300,
  width        = 2400,
  height       = 2400,
  plot_type = "volcano_plot"
)

p_volcano
```

# PRS-protein associations - with BMI adjustment

```{r protein-associations-bmi-adj}
protein_results_bmi_adj <- run_lm_grid(
  df           = data_protein,
  outcomes     = PROTEINS,
  predictors   = "prs",
  covariates   = c(COVARIATES, "z_fasting_time", "z_bmi"),
  idcol        = "eid",
  conf_level   = CONF_LEVEL,
  trait_map    = panel_lookup,
  trans_map    = setNames(character(0), character(0)),
  panel_lookup = panel_lookup,
  model_filter = "linear"
) %>% dplyr::mutate(pvalue_fdr = stats::p.adjust(pvalue, method = "fdr"))
```

```{r volcano-plot-bmi-adj, fig.width=16, fig.height=4}
p_volcano_bmi_adj <- plot_volcano_prs_proteins(
  results_df   = protein_results_bmi_adj,
  prs_name_map = c(prs = "PRS"),
  cov_formula  = c(COVARIATES, "z_fasting_time", "z_bmi"),
  top_n        = 10,
  p_nominal    = 0.05,
  fdr_cutoff   = 0.05,
  x_step       = 0.2,
  outputpath   = OUTPUT_FIGURES,
  save         = TRUE,
  dpi          = 300,
  width        = 2400,
  height       = 2400,
  plot_type = "volcano_plot_bmi_adj"
)

p_volcano_bmi_adj
```


## Export results to Excel

Two separate xlsx files are produced:
- `results_*.xlsx`  — models without BMI adjustment
- `results_bmi_adjusted_*.xlsx` — models with z_bmi as additional covariate

```{r export-results}
if (!dir.exists(OUTPUT_RESULTS)) dir.create(OUTPUT_RESULTS, recursive = TRUE)

write_workbook <- function(sheet_list, filepath) {
  wb <- openxlsx::createWorkbook()
  for (sheet_name in names(sheet_list)) {
    openxlsx::addWorksheet(wb, sheet_name)
    openxlsx::writeData(wb, sheet_name, sheet_list[[sheet_name]])
  }
  openxlsx::saveWorkbook(wb, filepath, overwrite = TRUE)
  message("saved: ", filepath)
}

# ── File 1: without BMI adjustment ────────────────────────────────────────────
out_no_bmi <- file.path(OUTPUT_RESULTS, make_table_filename("results"))

write_workbook(
  list(
    population_summary      = population_summary,
    linear_regression_prs   = linear_results,
    logistic_regression_prs = logistic_results,
    olink_proteins_prs      = protein_results
  ),
  out_no_bmi
)

# ── File 2: with BMI adjustment ────────────────────────────────────────────────
out_bmi <- file.path(OUTPUT_RESULTS, make_table_filename("results_bmi_adj"))

write_workbook(
  list(
    population_summary      = population_summary,
    linear_regression_prs   = linear_results_bmi_adj,
    logistic_regression_prs = logistic_results_bmi_adj,
    olink_protein_prs = protein_results_bmi_adj
  ),
  out_bmi
)
```


```{r}
p_dentures_prop <- plot_binary_by_prs_decile(
  data = data,
  position_type = "fill",
  var = "has_dentures",
  output_dir = OUTPUT_FIGURES,
  datatype = "phenotype",
  plot_type = "dentures_prs_deciles_fill",

      width = 2400,
      height = 2400
)

p_dentures_prop
```

```{r}
PHENOTYPES_OF_INTEREST <- c("CST5", "CA12", "CST1", "CGREF1", "GHRL")

protein_plots <- purrr::map(
  PHENOTYPES_OF_INTEREST,
  function(protein_name) {
    
    plot_protein_by_prs_decile(
      data = data_protein,
      protein = protein_name,
      output_dir = OUTPUT_FIGURES,
      datatype = "protein",
      plot_type = paste0(
        protein_name,
        "_prs_deciles_median_iqr"
      ),
      LOWER = -1,
      UPPER = 1,
      width = 2400,
      height = 2400
    )
  }
)

names(protein_plots) <- PHENOTYPES_OF_INTEREST
```
